---
title: "Spread of a pollutant in Lake Como"
output: rmarkdown::html_vignette
description: >
  This document explain the basic usage of femR to solve PDEs in R. 
vignette: >
  %\VignetteIndexEntry{Spread of a pollutant in Lake Como}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align= "center",
                                   collapse = TRUE,
                                   comment = "#>", 
                                   fig.width= 6,fig.height= 4.5,
                                   message = FALSE,
                                   warning = FALSE)
                                   
library(sf, quietly = T)
library(mapview, quietly = T)
library(femR, quietly = T)
library(rmapshaper, quietly = T)
```

Lake Como, nestled in the Lombardy region, is recognized as a crucial freshwater ecosystem esteemed for its biodiversity and ecological significance. Let assume that recent environmental studies suggests the potential presence and spreading of a concerning pollutant within the lake's waters. The hypothetical emergence of this pollutant raises significant concerns regarding the potential threat it may pose to the ecological balance of the aquatic environment and prompts considerations about its long-term environmental stability. The hypothetical presence of this pollutant prompts a thorough examination to understand its hypothetical nature, source, and potential implications on Lake Como's delicate ecosystem.
Assume that several monitoring stations have been strategically established around Lake Como to track and measure pollutant concentrations in its waters. These stations serve as pivotal tools for continuous surveillance, providing crucial data to assess the levels and potential impacts of pollutants on the lake's ecosystem.
In the following interactive figure, has been reported the comprehensive map of Lake Como includes georeferenced black points denoting the strategic placement of monitoring stations across its expanse. 
The red points denote potential 'hotspots' where abnormal concentrations of the pollutant have been recorded. 
```{r, echo=FALSE}
lake_bd <- read_sf("data/deims_sites_boundariesPolygon.shp")
lake_bd_simp <- rmapshaper::ms_simplify(lake_bd, keep= 0.5, 
                                        keep_shapes=T)
st_crs(lake_bd_simp) <- 4326
domain <- Domain(st_geometry(lake_bd_simp))
mesh <- build_mesh(domain, maximum_area = 0.25e-5, minimum_angle = 20)
mesh_sf <- st_as_sfc(mesh)
set.seed(0)
num_stations <- 20
boundary_points <- which(mesh$get_boundary() == 1)
boundary_id <- sample(x = 1:length(boundary_points),size = num_stations)
stations <- lapply(as.list(as.data.frame(t(mesh$get_nodes()[boundary_id,]))),
                   st_point)

stations <- st_sfc(stations, crs=4326)
station_name <- as.character(1:num_stations)
stations <- st_as_sf(data.frame(station_name=station_name),
                     geometry = stations)

sources <- st_as_sf(data.frame(station_name=station_name[c(16,9)]),
                    geometry=st_geometry(stations)[c(16,9)])
map.type <- "Esri.WorldTopoMap"
mappa_completa <-  mapview(stations, legend=F, cex=5, col.region="black",
        map.type=map.type) + 
  mapview(sources, legend=F, cex=5, col.region="red") +
  mapview(lake_bd_simp, col.regions = "white", lwd = 2, 
          alpha.regions=0.2,legend=F)
mappa_completa

mappa_mesh <- mapview(mesh_sf, col.region="white",alpha.regions=0.2,legend=F,  alpha=0.2,
        lwd=0.4,
        map.type=map.type) + 
  mapview(lake_bd_simp, col.regions = "white", lwd = 2, 
          alpha.regions=0.2,legend=F)

beta <- lapply(st_geometry(stations)[c(9,8)], as.numeric)
beta <- matrix(c(beta[[1]][1]-beta[[2]][1],
                 beta[[1]][2]-beta[[2]][2]), nrow=2)*10^2

mu = 0.01
Q <- 5
```

To model the spreads of the pollutant, one should consider also the presence of a tributary in the northern region of the lake. Hence, a transport term has been identified, signifying the movement of the pollutant carried by the inflow towards other areas within Lake Como. Moreover, non-homogeneous Dirichlet boundary condition are prescribed on boundary regions near the locations of the stations which recorded anomalies concerning the concentration of the pollutant. On the boundary regions unaffected by the anomalies, homogeneous Dirichlet boundary conditions are prescribed.

## Steady-state solution 
In this section, we consider the solution of the steady-state problem at hand.
Denote with $\Omega$ our domain of interest. Assume that the concentration of pollutant, denoted by $u$, spreads on $\Omega$ according to the following PDE:

$$
\begin{cases}
-\mu \Delta u + \boldsymbol{\beta} \cdot \nabla u = 0 \qquad & in \ \Omega \\
 u = g|_{\partial \Omega} \qquad         & on \ \partial \Omega,
\end{cases}
$$
where $\mu \in \mathbb{R}$, $\boldsymbol{\beta} \in \mathbb{R}^2$ are the diffusion coefficient and the transport coefficient, respectively. The function $g$ exploited to impose the boundary conditions is the sum of two Gaussian-like functions centered at the locations of the stations which recorded anomalous concentrations of pollutant.

The following sections explain how to solve the problem exploiting `femR` package. 
```{r, eval=FALSE}
library(sf, quietly = T)
library(mapview, quietly = T)
library(femR, quietly = T)
library(rmapshaper, quietly = T)

lake_bd <- read_sf("data/deims_sites_boundariesPolygon.shp")
lake_bd_simp <- rmapshaper::ms_simplify(lake_bd, keep= 0.5, 
                                        keep_shapes=T)
st_crs(lake_bd_simp) <- 4326

```

The `femR` package implements utilities to read a `sfc` (Simple Features Collection) object, such as the boundary of the Lake, and construct meshes by specifying the `maximum_area` of each triangle and the `minimum_angle` between adjacent sides of the triangles.
```{r}
# create the physical domain
domain <- Domain(st_geometry(lake_bd_simp)) 
# create the mesh
mesh <- build_mesh(domain, maximum_area = 0.25e-5, minimum_angle = 20)
```
In addition, the package provides utilities to convert the newly created mesh into an `sfc` object. For example, it allows us to visualize the mesh object on an interactive map using the `mapview` package, as shown in the following code chunk.
```{r}
mesh_sf <- st_as_sfc(mesh)

map.type <- "Esri.WorldTopoMap"
mapview(lake_bd_simp, col.regions = "white", lwd = 2, 
          alpha.regions=0.2,legend=F,map.type=map.type) +
mapview(mesh_sf, col.region="white",alpha.regions=0.2,legend=F,  alpha=0.4,
        lwd=0.4, map.type=map.type) 
```
Once, the mesh has been built, the `FunctionSpace`, to which the solution of the problem belongs, has to be defined. By the default, first order finite elements are considered when `Vh` is build. Then, we initialize `u`, the `Function` object, solution of the PDEs, as an instance of the object `Vh`.
```{r}
Vh <- FunctionSpace(mesh)
u <- Function(Vh)
```

Then, we define the differential operator in its strong formulation as follows:
```{r}
mu <- 0.001
beta <- c(-4.091349, -4.493403)/20
Lu <- -mu*laplace(u) + dot(beta, grad(u))
```

Either the forcing term of the differential problem and the boundary condition can be defined as simple R `function`s. You can use also numbers in case of either constant forcing term or constant boundary conditions. 
```{r}
# Dirichlet boundary conditions 
g <- function(points){
  res <- matrix(0, nrow=nrow(points), ncol=1)
  pts_list <- lapply(as.list(as.data.frame(t(points))), st_point)
  dists2 <- st_distance(st_geometry(sources),
                        st_sfc(pts_list, crs=st_crs(sources))) 
  dists2 <- matrix(as.numeric(dists2), 
                   nrow=length(st_geometry(sources)), ncol=nrow(points))/(2.5*10^3)
  for(i in 1:length(sources))
    res <- res + Q*exp(-dists2[i,]) 
  return(res)
}
```
Finally, we build a `pde` object passing the differential operator `L`, as first parameter and the forcing term `f`, as second parameter:
```{r, results="hide"}
# build PDE object
pde <- Pde(Lu, 0.)
# set boundary conditions
pde$set_boundary_condition(fun= g, type= "dirichlet")
```
We can compute the discrete solution of the problem calling the `solve` method:
```{r}
pde$solve()
```
Finally, it is possible to visualize the computed solution on an interactive map using the `mapview` package as the following code chunk shows.
```{r}
coeff <- apply(mesh$get_elements(), MARGIN=1, FUN=
                   function(edge){
                     mean(u$coeff[edge])})
U <- st_sf(data.frame(coeff = coeff), geometry= mesh_sf)
mapview(U, alpha.regions=0.8, alpha = 0, map.type=map.type)
```

```{r,eval=FALSE, include=FALSE}
library(webshot2)
library(leaflet)
cntr_crds <- c(mean(mesh$get_nodes()[, 1]),
               mean(mesh$get_nodes()[, 2]))

solution <- mapview(U, alpha.regions=0.8,  legend=F,
                    alpha = 0,                
                    map.type=map.type)
folder.name <- paste0(getwd(),"/../tests/")

html_fl =  paste0(folder.name,"mappa_mesh.html")
png_fl = paste0(folder.name, "mappa_mesh.png")
mappa_mesh@map <- mappa_mesh@map %>% setView(cntr_crds[1], cntr_crds[2], zoom = 10)
mapshot2(mappa_mesh, url = html_fl, file = png_fl, delay=10, 
         cliprect = c(190,135, 600,500), zoom=1.5 )


html_fl = paste0(folder.name, "mappa_completa.html")
png_fl = paste0(folder.name, "mappa_completa.png")
mappa_completa@map <- mappa_completa@map %>% setView(cntr_crds[1], cntr_crds[2], zoom = 10)
mapshot2(mappa_completa, url = html_fl, file = png_fl, delay=10, 
         cliprect = c(190,135, 600,500), zoom=1.5)


html_fl = paste0(folder.name, "solution_zoom.html")
png_fl = paste0(folder.name, "solution_zoom.png")
solution@map <- solution@map %>% setView(as.numeric(st_geometry(stations)[17][[1]])[1], 
                                         as.numeric(st_geometry(stations)[17][[1]])[2], zoom = 10.6)
mapshot2(solution, url = html_fl, file = png_fl, delay=10, 
         cliprect = c(190,135, 600,500), zoom=1.5)

html_fl = paste0(folder.name, "solution.html")
png_fl = paste0(folder.name, "solution.png")
solution@map <- solution@map %>% setView(as.numeric(st_geometry(stations)[17][[1]])[1], 
                                         as.numeric(st_geometry(stations)[17][[1]])[2], zoom = 10)
mapshot2(solution, url = html_fl, file = png_fl, delay=10, 
         cliprect = c(190,145, 600,500), zoom=1.5)

```

## Time-dependent solution

Now, we aim to estimate the evolution of pollutant concentrations in Lake Como. We consider the same Gaussian-like function $g$ of the previous section, appropriately modified to account the presence of time. We leverage this function to set both the initial condition for the pollutant concentration and to impose the boundary conditions. Hence, the partial differential equation (PDE) that models the evolution of pollutant concentration in Lake Como reads as follows:

$$
\begin{cases}
\frac{\partial u}{\partial t} -\mu \Delta u + \boldsymbol{\beta} \cdot \nabla u = 0 \qquad & in \ \Omega \times (0,T) \\
 u = g            & in \ \Omega, \ t=0  \\
 u = g|_{\partial \Omega} \qquad         & on \ \partial \Omega \times (0,T),

\end{cases}
$$

The following window wraps all the steps needed to estimate the evolution of the pollutant concentration relying on `femR` package.
```{r, results='hide'}
time_interval <- c(0,0.3)
times <- seq(time_interval[1], time_interval[2], length.out=50)
# function space
Vh <- FunctionSpace(mesh %X% times)
# PDE solution
u <- Function(Vh)
# differential operator
Lu <- dt(u) - mu*laplace(u) + dot(beta, grad(u))
# dirichlet bc
g <- function(points, times){
  res <- matrix(0, nrow=nrow(points), ncol=length(times))
  pts_list <- lapply(as.list(as.data.frame(t(points))), st_point)
  dists2 <- st_distance(st_geometry(sources),
                        st_sfc(pts_list, crs=st_crs(sources))) 
  dists2 <- matrix(as.numeric(dists2), 
                   nrow=length(st_geometry(sources)), ncol=nrow(points))/(2.5*10^3)
  for(i in 1:length(sources)){
      res[,1:length(times)] <- res[,1:length(times)] + Q*exp(-dists2[i,])
  }
  return(res)
}
# initial condition
u0 <- g(mesh$get_nodes(), times[1])
# build PDE object
pde <- Pde(Lu, 0.)
# set initial conditions
pde$set_initial_condition(u0)
# set boundary conditions
pde$set_boundary_condition(fun= g, type= "dirichlet")
# compute the discrete solution
pde$solve()
# plot 
plot(u)
```
