---
title: "Solving a parabolic problem"
output: rmarkdown::html_vignette
description: >
  This document explains how to solve a parabolic problem using femR package. 
vignette: >
  %\VignetteIndexEntry{Solving a parabolic problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width= 7., 
                                   fig.height= 5., 
                                   fig.align= "center",
                                   collapse = TRUE,
                                   comment = "#>")

library(femR)
options(warn = -1)
```

Let $\Omega$ be the unit square $[0,1]^2$.
Consider the following problem  defined over $\Omega \times [0,T]$:

$$
\begin{cases}
\frac{\partial u}{\partial t}-\Delta u  = f  &in \ \Omega \times [0,T] \\
      \quad u = u_0  &in \ \Omega, \ t=0 \\ 
      \quad u = 0     &on \ \partial \Omega,\  t>0
\end{cases}
$$


where $u_0 = \sin( 2\pi x)\sin(2\pi y)$ is the initial condition, $f(x,y) = 8\pi^2 \sin( 2\pi x)\sin(2\pi y) e^{-t}$ is the forcing term and $\partial \Omega$ is the boundary of $\Omega$ where we have prescribed homogeneous Dirichelet boundary condition for all $t$ grater than $0$.
The exact solution of the previous problem is $u_{ex}(x,y) = \sin(2\pi x) \sin(2 \pi y) e^{-t}$ whose trend is shown in the following.

<!-- ```{r, echo=FALSE} -->
<!-- exact_solution <- function(points,t){ -->
<!--   res <- matrix(0, nrow=nrow(points), ncol=length(times)) -->
<!--   for( t in 1:length(times)){ -->
<!--     res[,t] = sin( 2.* pi * points[,1]) * sin(2.*pi* points[,2]) * exp(-times[t]) -->
<!--   } -->
<!--   return(res) -->
<!-- } -->

<!-- t0 = 0. -->
<!-- t_max = 1 -->
<!-- dT = 1e-2 -->
<!-- M = (t_max - t0)/dT + 1 -->
<!-- times = seq(t0,t_max,length=M) -->

<!-- data("unit_square", package = "femR") -->
<!-- domain <- Mesh(unit_square) %X% times -->
<!-- eval_points <- unit_square$nodes -->
<!-- data_plot <- data.frame(X=rep(eval_points[,1], times=length(times)), Y=rep(eval_points[,2], times=length(times)), -->
<!--                         coeff=as.vector(exact_solution(eval_points, times)), -->
<!--                         t=rep(times, times=nrow(eval_points))) -->
<!-- I=(domain$elements()[,1]); J=(domain$elements()[,2]); K=(domain$elements()[,3])  -->
<!-- limits = c(min(data_plot$coeff), max(data_plot$coeff)) -->

<!-- fig_exact <- plot_ly(data_plot, type="contour", x=~X, y=~Y, z=~coeff, intensity=~coeff,color = ~coeff, -->
<!--                      frame =~t, -->
<!--                  i = I, j = J, k = K, zmin=limits[1], zmax=limits[2], -->
<!--           contours=list(showlabels = TRUE), -->
<!--           colorbar=list(title="")) %>% -->
<!--    layout(xaxis = list(title = "", showgrid=F, zeroline=F, ticks="", showticklabels=F), -->
<!--           yaxis = list(title = "", showgrid=F, zeroline=F, ticks="", showticklabels=F)) %>% -->
<!--       animation_opts(frame=5) %>%  -->
<!--       animation_slider(currentvalue = list(prefix ="t = ")) -->
<!-- ``` -->

<!-- ```{r, echo=FALSE} -->
<!-- fig_exact -->
<!-- ``` -->

```{r}
## load domain data
data("unit_square", package="femR")

## time steps
t0 = 0.
t_max = 1
dT = 1e-2
M = (t_max - t0)/dT + 1
times = seq(t0,t_max,length=M)

## Spatio-temporal domain
domain = Mesh(unit_square) %X% times

## create Functional Space
Vh <- FunctionSpace(domain)

## define differential operator in its strong formulation
u <- Function(Vh)

## define differential operator in its strong formulation
Lu <- dt(u) -laplace(u)

## forcing term
f <- function(points,times){
  res <- matrix(0, nrow=nrow(points), ncol=length(times))
  for( t in 1:length(times)){
    res[,t] = (8*pi^2 -1) * sin( 2.* pi * points[,1]) * sin(2.*pi* points[,2]) * exp(-times[t])
  }
  return(res) 
}
## Dirichlet BC
dirichletBC <- function(points, times){
  return(matrix(0, nrow=nrow(points), ncol=length(times)))
}

## initial condition 
initialCondition <- function(points){
 return(sin( 2.* pi * points[,1]) * sin(2.*pi* points[,2]))
}
  
## create pde
pde <- Pde(Lu, f)

# set initial condition and dirichlet BC
pde$set_initialCondition(initialCondition)
pde$set_dirichletBC(dirichletBC)

## solve problem
pde$solve()

## plot solution
contour(u)
```
